Overlay databanks unlock data-driven analyses of biomolecules for all

### 2023-05-10 main.tex
Update title, introduction and discussion based on feedback from a journal

\title{NMRlipids Databank makes data-driven analysis of biomembrane properties accessible for all}
%NMRlipids Databank: Making data-driven analyses of membrane properties accessible for all^
%Overlay Databank of Lipid Membrane Simulations Arising from Open Collaboration

@@

\begin{abstract}
Cellular membrane lipid composition is implicated in diseases and controls major biological functions, but membranes are difficult to study experimentally due to their intrinsic disorder and complex phase behaviour. Molecular dynamics (MD) simulations have been useful in understanding membrane systems, but they require significant computational resources and often suffer from inaccuracies in model parameters. Applications of data-driven and machine learning methods, currently revolutionising many fields, remain of limited use for membrane systems due to the lack of suitable training sets. Here we present the NMRlipids Databank---a community-driven, open-for-all database featuring programmatic access to quality-evaluated atom-resolution MD simulations of lipid bilayers. The NMRlipids Databank will benefit scientists in different disciplines by providing automatic ranking of simulations based on their quality against experiments, programmable interface for flexible implementation of data-driven and machine learning applications, and rapid access to simulation data through a graphical user interface. To demonstrate how it unlocks possibilities beyond current MD simulation studies, we analyzed the NMRlipids Databank to reveal how anisotropic diffusion of water and cholesterol flip-flop rates depend on membrane properties.

@@

Cellular membranes contain hundreds of different types of lipid molecules that regulate the membrane properties, morphology, and biological functions~\cite{vanmeer08,Lorent:2020a,Slatter:2016a}. Membrane lipid composition is implicated in diseases, such as cancer and neurodegenerative disorders, and therapeutics that affect membrane compositions are emerging~\cite{torres21}. However, biomembranes are often difficult to study experimentally, because they are complex mixtures of proteins and lipids in disordered fluid state with complicated phase behaviour at physiologicalbiological conditions. For those reasons, the detailed connections between complex lipid interactions and biological functions taking place in or around membranes remain poorly understood. Molecular dynamics (MD) simulations have been particularly useful in understanding membrane systems, although their accuracy has often been compromised by artefacts such as the quality of model parameters~\cite{antila22b,gupta22}. Presently, the accuracy of models is becoming increasingly important as researches are progressing from simulations of individual molecules to simulating whole organelles or even cells using interdisciplinary approaches~\cite{johnson15,thornburg22,gupta22}. Such systems exhibit intricate emergent behaviour making inaccuracies more difficult to detect, and accumulation of even modest errors may have a dramatic impact on the conclusions drawn.

In contrast to experimental structural biology, where standard protocols to share and quality-evaluate resolved biomolecular structures are firmly established by the Protein Data Bank (PDB)~\cite{montelione13}, equivalent best practices are yet to be defined for MD simulations. The importance of such approaches is widely recognized~\cite{feig99,tai04,silva06,abraham19,hildebrand19,hospital20,abriata20,espigares20} and data-sharing solutions are emerging for proteins in solution~\cite{meyer10,kamp10}, proteins in membranes~\cite{newport19,espigares20,leston22}, nucleic acids~\cite{hospital16}, nucleic acids and proteins~\cite{bekker20}, cyclodextrins~\cite{mixcoha16}, COVID-19--involved macromolecules~(\href{https://bioexcel-cv19.bsc.es}{bioexcel-cv19.bsc.es}). However, automatic quality evaluation of simulation data~\cite{meyer10,hospital16} and programmatic access are still rare. In particular, tools for automatic quality evaluation of membrane simulations, or training sets for machine learning models of membrane-containing systems, are not yet available. Recent advances using machine learning approaches utilizing publicly available databanks, for example to solve protein structures~\cite{jumper21}, emphasize the increasing importance of such resources.

Here we present the NMRlipids Databank --- a community-driven, open-for-all database featuring programmatic access to atom-resolution MD simulations of lipid bilayers. The programmatic access enables users to apply their data-driven approaches on the Databank, thus facilitating the creation of new tools for (and by) researchers in a wide range of fields covering academia and industry, from cell membrane biology and lipid nanoparticle formulations to computational chemistry and machine learning. As two usage examples of what is already possible, we demonstrate here (i) how data-driven analysis of water anisotropic diffusion in all the membrane systems available in the Databank can extend the scope of MD simulations to magnetic resonance imaging (MRI) and pharmacokinetics, and (ii) how the Databank allows its users to analyse rare phenomena that are beyond the scope of standard MD simulation investigations. Wider adaptation of the NMRlipids Databank will open even more possibilities. Furthermore, the Databank performs automatic quality evaluation of membrane simulations, which facilitates the selection of best-performing models for each given application and accelerates the development of simulation parameters and methodology.

While the NMRlipids Databank currently contains only lipid bilayer systems, its key elements that enable programmatic access to large-scale MD simulation data can be applied also for other molecules, such as disordered proteins or sugars.
%The powerful combination of an overlay databank structure and a community-wide open collaboration can be used to build databases that enable data-driven and machine learning applications also in other fields, where storage of raw data requires significant resources, best practices in the field are not yet defined, and incentives to share data do not exist---such as assignment of NMR spectra~\cite{klukowski22}. 
The powerful combination of an overlay databank structure and the community-wide open collaboration  described here is potentially useful for building databases that enable data-driven and machine learning applications also in other fields. This is especially true for research activities where storage of raw data requires significant resources, best practices in the field are not yet defined, and incentives to share data do not exist---such as assignment of NMR spectra~\cite{klukowski22}.

@@

\subsection{Applications of the NMRlipids Databank}
\subsubsection{Selecting simulation parameters using NMRlipids Databank: Best models for most abundant neutral membrane lipids}
%Quality of membrane simulations with different force fields have been evaluated against experimental data during parameterization and in separate comparison studies~\cite{botan15,ollila16,catte16,pluhackova16,perez17,leonard19,??}, but the lack of universal quality measure for membrane simulations complicates the estimations of reliability of simulations. 
%
%For example, OPLS3e parameters overcome CHARMM36 in structural quality for POPC but predicts overestimated ion binding~\cite{kurki22}. On the other hand, CHARMM36 gives the best description for lipid headgroup conformational ensembles~\cite{bacle21}, while GROMOS-CKP parameters best capture the membrane packing in POPS bilayers~\cite{antila22b}. 
%
To minimize the detrimental consequences of artificial MD simulation results for their applications, the quality of lipid bilayer MD simulations has to be carefully assessed~\cite{antila22b}. This can be done, for example, against the C--H bond order parameters from NMR spectroscopy~\cite{bacle21,Wur23} and the form factors from X-ray scattering~\cite{ollila16}, although it requires comparisons between large number of simulations, which is laborious even with collaborative approaches~\cite{botan15,catte16,antila19,bacle21}. To streamline this process, we have defined quantitative quality measures that enable automatic ranking of lipid bilayer simulations based on their quality against experiments. Conformational ensembles of individual lipid molecules are evaluated in the NMRlipids Databank by first calculating the probabilities for each C--H bond order parameter to locate within experimental error and then averaging the possibilities over different lipid segments ($P^{\mathrm{hg}}$, $P^{\mathrm{sn1}}$,$P^{\mathrm{sn2}}$, and $P^{\mathrm{total}}$). Furthermore, the quality against X-ray scattering experiments ($FF_q$) is estimated as the difference in the experimental and simulated locations of the first form factor minimum. These measures can be used to evaluate membrane dimensions, as the form factor minima locations and acyl chain C--H bond order parameters correlate with the membrane lateral packing and thickness (Figs.~\ref{fig:quality}G,~\ref{fig:QualityCorrelationsSI}, and~\ref{fig:sizedependence}). In addition, the ergodicity of conformational sampling of lipids is estimated by calculating $\tau_\mathrm{rel}$, the convergence time of the slowest principal component divided by the simulation length. Here we demonstrate how the automatic simulation-quality evaluation and the NMRlipids Databank-API enable rapid selection of the best models for simulations of membranes with the two most biologically-abundant neutral membrane lipids~\cite{vanmeer08}: POPC and POPE.

@@

The focus of biomolecular simulations is moving from studies of individual molecules to larger complexes and even whole cells and organelles~\cite{johnson15,thornburg22,gupta22}. Simultaneously, machine-learning-based models for predicting the behaviour of biomolecules and automatic approaches to parametrize models are emerging~\cite{jumper21,antila22b}. The resources delivered by the NMRlipids Databank will support the development in both of these directions. Automatic quality evaluation and ranking of simulations against experimental data enable the selection of best simulations for specific applications without laborious manual force field evaluation. This also streamlines automatic parametrization procedures for atomistic and coarse grained simulations by, for example, pinpointing typical failures of force fields and highlighting points of improvement. Such practises for fostering the accuracy of simulations are becoming increasingly important due the accumulation of small errors when complexity and size of simulated systems are increasing.

The open programmatic access to unprecedented amounts of MD simulation data through the NMRlipids Databank-API enables a wide range of novel data-driven analyses and machine learning applications. The analyses of cholesterol flip-flop events (Fig.~\ref{fig:flip-flops}) and water permeation through membranes (Fig.~\ref{fig:permeability}) demonstrate how a large amount of accessible simulation data in terms of quantity (\textit{e.g.}, simulation length and number of conformations) and content (\textit{e.g.}, lipid compositions and ion concentrations) enable analyses of rare phenomena that are beyond the current possibilities for a single research group. Such analyses also pave the way for applications of MD simulations in new fields, as demonstrated here by analysing an essential parameter in pharmacokinetic modeling and MRI imaging:\cite{nitsche19,topgaard20} the anisotropic diffusion of water in membrane systems (Fig.~\ref{fig:permeability}). Furthermore, the NMRlipids Databank-API delivers access to a training set data that paves the way for diverse machine learning applications to predict membrane properties. Such applications could be analogous to AlphaFold \cite{jumper21} and other tools~\cite{baek21,lin23} that predict protein structures from their sequence using artificial intelligence. These possibilities are particularly valuable for scientists who do not typically have access to large scale MD simulation data. Expected applications of the NMRlipids Databank in a wide range of fields are listed in Table~\ref{tab:applications}, yet the scope of such applications
%in biology, biotechnology, material science, biomolecular imaging and beyond
is expected to further widen with increasing amount of data in the Databank.

\begin{table}[t]
    \centering
    \begin{tabular}{p{5.0cm}  p{5.0cm}  p{4.0cm}}
    Type of application     & Practical examples & Target group \\
    \hline
    Analyses of rare phenomena               & Lipid flip-flops, water permeation (Figs.~\ref{fig:flip-flops} and~\ref{fig:permeability}) & Membrane scientists \\
    \\
    Correlations between membrane properties & 
    Membrane structural properties, water dynamics (Figs.~\ref{fig:quality} and~\ref{fig:permeability}) & 
    Membrane scientists \\
    \\
    Applications that are outside the typical scope of MD simulations & 
    Anisotropic water diffusion for pharmacokinetics and MRI imaging applications (Fig.~\ref{fig:permeability}) & 
    Scientists in fields where MD simulations are not usually applied \\
    \\
    Selection of the best simulation model for a specific application & 
    Best model for membranes with PC and PE lipids (Fig.~\ref{fig:quality}), lipid headgroup conformations~\cite{bacle21}, 
    packing of PS~\cite{antila22b} and PE (Fig.~\ref{fig:quality}) containing membranes &
    Scientists using MD simulations \\
    \\
    Guidance for force field development & 
    Improvements in ion binding to lipids~\cite{melcr18,melcr20} and lipid headgroup conformational ensembles~\cite{yu21,dickson22,grote20} &
    Scientists developing parameters for MD simulations \\
    \\
    Training and target data for the development of coarse grained models & 
    Optimizing parameters of coarse grained models against NMRlipids Databank, extracting continuum parameters for membranes &
    Scientists developing and using coarse grained MD simulations \\
    \\
    Training set for machine learning applicatons &
    Programmatic access to the data and results enables training of machine learning models for various applications, such as predictions of membrane properties from composition & Scientists building and using machine learning applications for biomolecules
    \end{tabular}
    \caption{Examples on applications of the NMRlipids Databank.}
    \label{tab:applications}
\end{table}

Potential benefits of sharing MD simulation data are widely recognized by different stakeholders from scientists~\cite{feig99,tai04,silva06,abraham19,hildebrand19,hospital20,abriata20,espigares20} to funders and journal staff, but data sharing with efficient tools for upcycling have been rare. The main barriers for such solutions have been the required commitment for the long term support of hardware and software, and the lack of incentives for researchers to share the data. The first issue is solved in the NMRlipids Databank by the overlay design, where the raw data is distributed to already publicly available decentralized locations, while the core of the Databank is composed only of the metadata stored in a version-controlled git repository with an open-access license. On the other hand, the open collaboration approach developed in the NMRlipids Project~\cite{botan15} creates incentives for sharing the data by offering authorship in published articles to the contributors. Advantages of such an approach are demonstrated here for membrane simulations, but the concept can be applied not only to other biomolecules, such as disordered proteins and membrane--protein systems, but also in other fields where similar barriers hinder the machine learning revolution, such as the assignment of NMR spectra~\cite{klukowski22}.

>>>>

\title{Overlay databanks unlock data-driven analyses of biomolecules for all}
%NMRlipids Databank makes data-driven analysis of biomembrane properties accessible for all}
%NMRlipids Databank: Making data-driven analyses of membrane properties accessible for all^
%Overlay Databank of Lipid Membrane Simulations Arising from Open Collaboration

@@

\begin{abstract}
Tools based on artificial intelligence (AI) are currently revolutionising many fields, yet their applications are often limited by the lack of suitable training data in programmatically accessible format. Here we propose an effective solution to make data scattered in various locations and formats accessible for data-driven and machine learning applications using the overlay databank format. To demonstrate the practical relevance of such approach, we present the NMRlipids Databank---a community-driven, open-for-all database featuring programmatic access to quality-evaluated atom-resolution molecular dynamics simulations of cellular membranes. Cellular membrane lipid composition is implicated in diseases and controls major biological functions, but membranes are difficult to study experimentally due to their intrinsic disorder and complex phase behaviour. While MD simulations have been useful in understanding membrane systems, they require significant computational resources and often suffer from inaccuracies in model parameters. Here, we demonstrate how programmable interface for flexible implementation of data-driven and machine learning applications, and rapid access to simulation data through a graphical user interface unlocks possibilities beyond current MD simulation and experimental studies to understand cellular membranes. The proposed overlay databank concept can be further applied to other biomolecules, as well as in other fields where similar barriers hinder the AI revolution.

%Cellular membrane lipid composition is implicated in diseases and controls major biological functions, but membranes are difficult to study experimentally due to their intrinsic disorder and complex phase behaviour. Molecular dynamics (MD) simulations have been useful in understanding membrane systems, but they require significant computational resources and often suffer from inaccuracies in model parameters. Applications of data-driven and machine learning methods, currently revolutionising many fields, remain of limited use for membrane systems due to the lack of suitable training sets. Here we present the NMRlipids Databank---a community-driven, open-for-all database featuring programmatic access to quality-evaluated atom-resolution MD simulations of lipid bilayers. The NMRlipids Databank will benefit scientists in different disciplines by providing automatic ranking of simulations based on their quality against experiments, programmable interface for flexible implementation of data-driven and machine learning applications, and rapid access to simulation data through a graphical user interface. To demonstrate how it unlocks possibilities beyond current MD simulation studies, we analyzed the NMRlipids Databank to reveal how anisotropic diffusion of water and cholesterol flip-flop rates depend on membrane properties.

@@

Tools based on artificial intelligence (AI) are currently revolutionising many fields, yet their success relies on training data that is available in programmatically accessible and standardized format~\cite{sever23}. For example in structural biology, Protein Data Bank (PDB)~\cite{montelione13} has enabled revolutionary tools that can predict protein structures with unprecedented accuracy~\cite{jumper21}. However, the lack of smart guidelines and community-consensus best-practices for data sharing are limiting the development of such databanks and AI applications in many fields~\cite{sever23}. This is the case in biomolecular modeling where vast amount of data is already available for proteins, lipids, nucleic acids and carbohydrates, but tools that enable applications of this data for data-driven applications are not yet available~\cite{tiemann23}. Here we propose a cost-effective solution to make data scattered in various locations and formats accessible for data-driven and machine learning applications using the overlay databank format. We demonstrate the practical relevance of such approach for understanding cellular membranes by incorporating available lipid bilayer simulations into the NMRlipids databank. Nevertheless, the basic principles of overlay databanks can be applied also to simulation data of other biomolecules that are becoming available in increasing amounts~\cite{tiemann23}, as well as to other fields where development of AI based tools are limited by the data access and best best community practices --- such as assignment of NMR spectra~\cite{klukowski22}. Incentives to share data can be further accelerated by combining overlay databanks with open collaboration approach developed by us~\cite{botan15}.

Cellular membranes contain hundreds of different types of lipid molecules that regulate the membrane properties, morphology, and biological functions~\cite{vanmeer08,Lorent:2020a,Slatter:2016a}. Membrane lipid composition is implicated in diseases, such as cancer and neurodegenerative disorders, and therapeutics that affect membrane compositions are emerging~\cite{torres21}. However, biomembranes are often difficult to study experimentally, because they are complex mixtures of proteins and lipids in disordered fluid state with complicated phase behaviour at biological conditions. Molecular dynamics (MD) simulations can be used to model biomembranes in detail, but computational cost for simulating all possible biological membrane compositions would be substantial. For those reasons, data-driven and machine learning based models that predict biomembrane properties will benefit wide range of fields covering academia and industry, from cell membrane biology to lipid nanoparticle formulations.
%For those reasons, the detailed connections between complex lipid interactions and biological functions taking place in or around membranes remain poorly understood. 
%Molecular dynamics (MD) simulations have been particularly useful in understanding membrane systems, although their accuracy has often been compromised by artefacts such as the quality of model parameters~\cite{antila22b,gupta22}. Presently, the accuracy of models is becoming increasingly important as researches are progressing from simulations of individual molecules to simulating whole organelles or even cells using interdisciplinary approaches~\cite{johnson15,thornburg22,gupta22}. Such systems exhibit intricate emergent behaviour making inaccuracies more difficult to detect, and accumulation of even modest errors may have a dramatic impact on the conclusions drawn.

%In contrast to experimental structural biology, where standard protocols to share and quality-evaluate resolved biomolecular structures are firmly established by the Protein Data Bank (PDB)~\cite{montelione13}, equivalent best practices are yet to be defined for MD simulations. The importance of such approaches is widely recognized~\cite{feig99,tai04,silva06,abraham19,hildebrand19,hospital20,abriata20,espigares20} and data-sharing solutions are emerging for proteins in solution~\cite{meyer10,kamp10}, proteins in membranes~\cite{newport19,espigares20,leston22}, nucleic acids~\cite{hospital16}, nucleic acids and proteins~\cite{bekker20}, cyclodextrins~\cite{mixcoha16}, COVID-19--involved macromolecules~(\href{https://bioexcel-cv19.bsc.es}{bioexcel-cv19.bsc.es}). However, automatic quality evaluation of simulation data~\cite{meyer10,hospital16} and programmatic access are still rare. In particular, tools for automatic quality evaluation of membrane simulations, or training sets for machine learning models of membrane-containing systems, are not yet available. Recent advances using machine learning approaches utilizing publicly available databanks, for example to solve protein structures~\cite{jumper21}, emphasize the increasing importance of such resources.

Here we present the NMRlipids Databank --- a community-driven, open-for-all database featuring programmatic access to atom-resolution MD simulations of lipid bilayers. To demonstrate its advances to available approaches, we show how rare phenomena that are beyond the scope of standard MD simulation investigations can be analyzed from the Databank. To demonstrate the immediate relevance of NMRlipids databank in extending the scope of MD simulations to new fields, we  use data-driven approach to analyse how anisotropic diffusion of water depends on membrane properties. This benefits understanding of magnetic resonance imaging (MRI)~\cite{topgaard20} and pharmacokinetics~\cite{nitsche19} where MD simulations are rarely applied.
%Wider adaptation of the NMRlipids Databank will open even more possibilities. 
Furthermore, the Databank performs automatic quality evaluation of membrane simulations, which facilitates the selection of best-performing models for each given application and accelerates the development of simulation parameters and methodology. The overlay databank and open collaboration approaches facilitate programmatic access and collection of community contributed data also in other fields than membrane simulations. For example, substantial biomolecular MD simulation data is already available but in scattered locations and formats~\cite{tiemann23}. The technical advances presented here will immediately benefit in making this data programmatically accessible. Furthermore, the overlay databank configuration will benefit broad range of fields where access to training data is bottle neck for building AI based tools.
%NMRlipids databanksolutions presented as well as broader communities that 
%enables users to apply their data-driven approaches on the Databank, thus facilitating the creation of new tools for (and by) researchers in a .
% to computational chemistry and machine learning.

%While the NMRlipids Databank currently contains only lipid bilayer systems, its key elements that enable programmatic access to large-scale MD simulation data can be applied also for other molecules, such as disordered proteins or sugars.
%%The powerful combination of an overlay databank structure and a community-wide open collaboration can be used to build databases that enable data-driven and machine learning applications also in other fields, where storage of raw data requires significant resources, best practices in the field are not yet defined, and incentives to share data do not exist---such as assignment of NMR spectra~\cite{klukowski22}. 
%The powerful combination of an overlay databank structure and the community-wide open collaboration  described here is potentially useful for building databases that enable data-driven and machine learning applications also in other fields. This is especially true for research activities where storage of raw data requires significant resources, best practices in the field are not yet defined, and incentives to share data do not exist---such as assignment of NMR spectra~\cite{klukowski22}.

@@

\subsection{Applications of the NMRlipids Databank}
\subsubsection{Selecting simulation parameters using NMRlipids Databank: Best models for most abundant neutral membrane lipids}
%Quality of membrane simulations with different force fields have been evaluated against experimental data during parameterization and in separate comparison studies~\cite{botan15,ollila16,catte16,pluhackova16,perez17,leonard19,??}, but the lack of universal quality measure for membrane simulations complicates the estimations of reliability of simulations. 
%
%For example, OPLS3e parameters overcome CHARMM36 in structural quality for POPC but predicts overestimated ion binding~\cite{kurki22}. On the other hand, CHARMM36 gives the best description for lipid headgroup conformational ensembles~\cite{bacle21}, while GROMOS-CKP parameters best capture the membrane packing in POPS bilayers~\cite{antila22b}. 
%
Molecular dynamics (MD) simulations have been particularly useful in understanding membrane systems, although their accuracy has often been compromised by artefacts such as the quality of model parameters~\cite{antila22b,gupta22}. Presently, the accuracy of models is becoming increasingly important as researches are progressing from simulations of individual molecules to simulating whole organelles or even cells using interdisciplinary approaches~\cite{johnson15,thornburg22,gupta22}. Such systems exhibit intricate emergent behaviour making inaccuracies more difficult to detect, and accumulation of even modest errors may have a dramatic impact on the conclusions drawn.
To minimize the detrimental consequences of artificial MD simulation results for their applications, the quality of lipid bilayer MD simulations has to be carefully assessed~\cite{antila22b}. This can be done, for example, against the C--H bond order parameters from NMR spectroscopy~\cite{bacle21,Wur23} and the form factors from X-ray scattering~\cite{ollila16}, although it requires comparisons between large number of simulations, which is laborious even with collaborative approaches~\cite{botan15,catte16,antila19,bacle21}. To streamline this process, we have defined quantitative quality measures that enable automatic ranking of lipid bilayer simulations based on their quality against experiments. Conformational ensembles of individual lipid molecules are evaluated in the NMRlipids Databank by first calculating the probabilities for each C--H bond order parameter to locate within experimental error and then averaging the possibilities over different lipid segments ($P^{\mathrm{hg}}$, $P^{\mathrm{sn1}}$,$P^{\mathrm{sn2}}$, and $P^{\mathrm{total}}$). Furthermore, the quality against X-ray scattering experiments ($FF_q$) is estimated as the difference in the experimental and simulated locations of the first form factor minimum. These measures can be used to evaluate membrane dimensions, as the form factor minima locations and acyl chain C--H bond order parameters correlate with the membrane lateral packing and thickness (Figs.~\ref{fig:quality}G,~\ref{fig:QualityCorrelationsSI}, and~\ref{fig:sizedependence}). In addition, the ergodicity of conformational sampling of lipids is estimated by calculating $\tau_\mathrm{rel}$, the convergence time of the slowest principal component divided by the simulation length. Here we demonstrate how the automatic simulation-quality evaluation and the NMRlipids Databank-API enable rapid selection of the best models for simulations of membranes with the two most biologically-abundant neutral membrane lipids~\cite{vanmeer08}: POPC and POPE.

@@

Sharing of biomolecular MD simulation and other data is becoming increasingly important in the age of big data and artificial intelligence~\cite{sever23,tiemann23}. 
%The main barriers for such solutions have been the required commitment for the long term support of hardware and software, and the lack of incentives for researchers to share the data. 
Besides the data itself, also programmatic access is a necessary requirement for data-driven and machine learning applications. This is particularly challenging when field specific smart guidelines and community-consensus best-practices have not yet been defined, which is the case in biomolecular simulations~\cite{tiemann23}. The NMRlipids Databank demonstrates how these issues can be solved by the overlay databank design where the raw data is distributed to already publicly available decentralized locations, while the core of the Databank is composed only of the metadata stored in a version-controlled git repository with an open-access license. On the other hand, the open collaboration approach developed in the NMRlipids Project~\cite{botan15} creates incentives for sharing the data by offering authorship in published articles to the contributors. Advantages of such an approach are demonstrated here for membrane simulations, yet the concept can be applied to other biomolecules, %such as disordered proteins and membrane--protein systems, 
as well as in other fields where similar barriers hinder the AI revolution, such as the assignment of NMR spectra~\cite{klukowski22}.

The NMRlipids Databank-API delivers programmatic access to MD simulation data that can be used as training set for diverse data-driven and machine learning applications that predict membrane properties. Such applications could be analogous to AlphaFold \cite{jumper21} and other tools~\cite{baek21,lin23} that predict protein structures from their sequence using artificial intelligence.
%The open programmatic access to unprecedented amounts of MD simulation data through the NMRlipids Databank-API enables a wide range of novel data-driven analyses and machine learning applications. 
The analyses of cholesterol flip-flop events (Fig.~\ref{fig:flip-flops}) and water permeation through membranes (Fig.~\ref{fig:permeability}) demonstrate how a large amount of accessible simulation data in terms of quantity (\textit{e.g.}, simulation length and number of conformations) and content (\textit{e.g.}, lipid compositions and ion concentrations) enable analyses of rare phenomena that are beyond the current possibilities for a single research group. Such analyses also pave the way for applications of MD simulations in new fields, as demonstrated here by analysing an essential parameter in pharmacokinetic modeling and MRI imaging:\cite{nitsche19,topgaard20} the anisotropic diffusion of water in membrane systems (Fig.~\ref{fig:permeability}). 
These possibilities are particularly valuable for scientists who do not typically have access to large scale MD simulation data.

The focus of biomolecular simulations is moving from studies of individual molecules to larger complexes and even whole cells and organelles~\cite{johnson15,thornburg22,gupta22}. Simultaneously, machine-learning-based models for predicting the behaviour of biomolecules and automatic approaches to parametrize models are emerging~\cite{jumper21,antila22b}. The resources delivered by the NMRlipids Databank will support the development in both of these directions. Automatic quality evaluation and ranking of simulations against experimental data enable the selection of best simulations for specific applications without laborious manual force field evaluation. This also streamlines automatic parametrization procedures for atomistic and coarse grained simulations by, for example, pinpointing typical failures of force fields and highlighting points of improvement. Such practises for fostering the accuracy of simulations are becoming increasingly important due the accumulation of small errors when complexity and size of simulated systems are increasing. Examples of impact of NMRlipids and other overlay Databanks in different disciplines are listed in Table~\ref{tab:applications}, yet the scope of applications is expected to further widen with increasing amount of publicly shared data.

\begin{table}[t]
    \centering
    \begin{tabular}{p{3.0cm}  p{4.0cm}  p{7.0cm}}
     Target group  &  Outcomes     & Practical examples \\
    \hline
    Computational and data scientists & Databanks with programmatic access, training sets for machine learning applications, automatic optimization of simulation models & Programmatic access to all publicly available MD simulation trajectories, machine learning models predicting properties of complex biomolecular assemblies, optimizing force field and model parameters from atomistic to continuum scale \\

    Biomedical scientists & Applications of biomolecular modeling in new fields, coupling omics to biomolecular structure and dynamics & Anisotropic water diffusion for pharmacokinetics and MRI imaging applications (Fig.~\ref{fig:permeability}), properties of complex cellular structures based on composition analyses from omics \\

    Material scientists & Predictions of complex biomolecular material properties from data-driven and machine learning models & Optimising composition of lipid nanoparticle formulations for desired properties, predictions of bioinspired material properties \\

    Biophysicists &  Novel analyses and predictions for properties of biomolecular assemblies and their compositions  & Analyses of rare phenomena (for example, Lipid flip-flops and water permeation in Figs.~\ref{fig:flip-flops} and~\ref{fig:permeability}), data-driven analyzes and machine learning models  predicting correlations between properties and compositions of biomolecular assemblies (Figs.~\ref{fig:quality} and~\ref{fig:permeability}) \\

    Students and teachers & Graphical and programmatic access to biomolecular structure and dynamics & Illustration of disorder and dynamics in biomolecules, example data for bioinformatics and data-science courses \\

    Reviewers and publishers & Tool to facilitate FAIRness of data, transparency and reproducibility & All publicly available data can be included in an overlay databank irrespectively on its location \\

    \end{tabular}
    \caption{Examples of impact of NMRlipids and other overlay databanks in different disciplines.}
    \label{tab:applications}
\end{table}

%\begin{table}[t]
%    \centering
%    \begin{tabular}{p{5.0cm}  p{5.0cm}  p{4.0cm}}
%    Type of application     & Practical examples & Target group \\
%    \hline
%    Analyses of rare phenomena               & Lipid flip-flops, water permeation (Figs.~\ref{fig:flip-flops} and~\ref{fig:permeability}) & Membrane scientists \\
%    \\
%    Correlations between membrane properties & 
%    Membrane structural properties, water dynamics (Figs.~\ref{fig:quality} and~\ref{fig:permeability}) %& 
%    Membrane scientists \\
%    \\
%    Applications that are outside the typical scope of MD simulations & 
%    Anisotropic water diffusion for pharmacokinetics and MRI imaging applications %(Fig.~\ref{fig:permeability}) & 
%    Scientists in fields where MD simulations are not usually applied \\
%    \\
%    Selection of the best simulation model for a specific application & 
%    Best model for membranes with PC and PE lipids (Fig.~\ref{fig:quality}), lipid headgroup %conformations~\cite{bacle21}, 
%    packing of PS~\cite{antila22b} and PE (Fig.~\ref{fig:quality}) containing membranes &
%    Scientists using MD simulations \\
%    \\
%    Guidance for force field development & 
%    Improvements in ion binding to lipids~\cite{melcr18,melcr20} and lipid headgroup conformational %ensembles~\cite{yu21,dickson22,grote20} &
%    Scientists developing parameters for MD simulations \\
%    \\
%    Training and target data for the development of coarse grained models & 
%    Optimizing parameters of coarse grained models against NMRlipids Databank, extracting continuum %parameters for membranes &
%    Scientists developing and using coarse grained MD simulations \\
%   \\
%    Training set for machine learning applicatons &
%    Programmatic access to the data and results enables training of machine learning models for various %applications, such as predictions of membrane properties from composition & Scientists building and %using machine learning applications for biomolecules
%    \end{tabular}
%    \caption{Examples on applications of the NMRlipids Databank.}
%    \label{tab:applications}
%\end{table}

%Potential benefits of sharing MD simulation data are widely recognized by different stakeholders from scientists~\cite{feig99,tai04,silva06,abraham19,hildebrand19,hospital20,abriata20,espigares20} to funders and journal staff, but data sharing with efficient tools for upcycling have been rare. The main barriers for such solutions have been the required commitment for the long term support of hardware and software, and the lack of incentives for researchers to share the data. The first issue is solved in the NMRlipids Databank by the overlay design, where the raw data is distributed to already publicly available decentralized locations, while the core of the Databank is composed only of the metadata stored in a version-controlled git repository with an open-access license. On the other hand, the open collaboration approach developed in the NMRlipids Project~\cite{botan15} creates incentives for sharing the data by offering authorship in published articles to the contributors. Advantages of such an approach are demonstrated here for membrane simulations, but the concept can be applied not only to other biomolecules, such as disordered proteins and membrane--protein systems, but also in other fields where similar barriers hinder the machine learning revolution, such as the assignment of NMR spectra~\cite{klukowski22}.
